---
title: "Lab 4: Continous Random Variables"
author: "kgteller" "Jxcai revised on 2024-03-05"
date: "2023-05-31"
output: 
  html_document:
    extra_dependencies: ["asmmath"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Continuous Random Variables

Today we will compare samples from a continuous random variable with different sample sizes to the theoretical distribution.

## The Normal Distribution

The probability distribution for a random variable X that is normally distributed is defined by:

$f(x)=\frac{1}{\sqrt{2 \pi} \sigma} e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}$ on $(-\infty,\infty)$ and $E(X)=\mu$ and $Var(x)=\sigma^2$

We can plot the standard normal distribution using dnorm() in R

```{r}


# dnorm(x, mean, sd) will give you the curve 
#(function) value at x. That curve is normal
# with mean and sd

# d means getting the function value
# norm means for normal distribution



x<-seq(-4,4,length=500)
y<-dnorm(x,0,1)
plot(x,y,type='l')
```

### Assignment 1

Now create a loop that plots the normal distribution with $\mu=0$ and $\sigma= 3,5,10$.

```{r}
mu<-0;

sigma<-c(3,5,10)

for (a in sigma) {
  x<-seq(-40,40,length=5000)
  y<-dnorm(x,mu,a)
  plot(x,y,type='l',ylim = c(0,0.14),main = paste('standard deviation=',a))
  
}

```

a.  Describe how the normal probability density function changes as the standard deviation increases.

    ```         
    As standard deviation increases, the spread becomes larger and the peak becomes higher.
    ```

b.  What is the maximum value of the probability density function for the normal distribution with mean 0 and standard deviation 1?

    ```{r}

    # first, plot the normal curve with mean 0 and sd 1.
    # Then, find out the maximum of the curve is 
    # obtained at which point
    # Then use dnorm(that point, 0, 1) to find 
    # the maximum y.

    #Plotting the curve
    mu <- 0 #setting mean to 0
    sigma <- 5 #setting standard deviation to 1
    x <- seq(-10, 10, length=5000) #Setting the range of the plot
    y <- dnorm(x, mu, sigma) #Getting the normal distribution
    plot(x,y,type='l') #Plotting

    #Finding the max point
    max_point <- x[which.max(y)] #Finding the max x 

    #Using dnorm to find the max y
    max_y <- dnorm(max_point, mu, sigma)

    #Printing the max
    cat("Max_y using dnorm: ", max_y)


    ```

    ```         
    0.3989415
    ```

c.  What is the maximum value of the probability density function for the normal distribution with mean 0 and standard deviation 5?

    ```         
    0.07978845
    ```

d.  If a random variable had a standard deviation of 0, what would that mean?

    ```{r}

    # Hint: Try to plot normal curve with mean 0 but sd as 0.1, 0.01, 0.001, and tell your observation.

    #Plotting the curve
    mu <- 0 #setting mean to 0
    sigma <- c(0.1, 0.01, 0.001, 0.0001) #creating list for standard deviation
    for(a in sigma) {
      x <- seq(-10, 10, length=5000) #Setting the range of the plot
      y <- dnorm(x, mu, a) #Getting the normal distribution
      plot(x,y,type='l',ylim = c(0,0.14),main = paste('standard deviation=',a)) #plotting
    }
    #Observation:
    # If a random variable had a standard deviation of 0, then the graph would be a straight line and there would be no "peak".
    ```

Now display five plots with $\sigma=1$ and $\mu=-3,-1,0,5,10$

```{r}
sigma = 1
mu = c(-3, -1, 0, 5, 10) 

#Looping through mu's
for (a in mu) {
  x <- seq(-15, 15, length=5000) #Setting the range of the plot
  y <- dnorm(x, a, sigma) #Getting the normal distribution
  plot(x,y,type='l',ylim = c(0,0.4),main = paste('standard deviation=',a)) #plotting
}



```

Describe what happens to the probability density function as the mean increases.

```         
As Mu (the mean) is increased, the peak of the graph is shifted torwards the right. Nothing else is modified.
```

**End of Assignment 1**

Now that we understand how the parameters effect the probability density function lets look at calculating probabilities from a normal distribution.

To find the value of the PDF at x we can use dnorm() as above. Therefor to find $f(2)$ from a normal distribution with $\mu=0$ and $\sigma=1$ we would do the following:

```{r}
ans<-round(dnorm(2,0,1),5)

print(paste('f(2)=',ans))
```

To find $P(X \leq x)$ we can use pnorm(). To find the $P(X \leq 2)$ for the standard Normal distribution we would do the following:

```{r}
ans<-round(pnorm(2,0,1),5)

print(paste('P(X<=2)=',ans))

```

Now in order to find $P(a \leq X \leq b)$ we can use pnorm and subtraction. To find $P(-2 \leq X \leq 2)$ for the standard normal distribution we would do the following:

```{r}


# To find P(A<=X<=B)
# USE
# pnorm(B,mean,sd)-pnorm(A,mean,sd)


ans<-round(pnorm(2,0,1)-pnorm(-2,0,1),5)

print(paste('P(-2<= X <=2)=',ans))
```

Finally to find a value q such that $P(X \leq q)=p$ we can use qnorm. This is a way to find percentiles for a normal distribution. To find the 95th percentile for the standard normal distribution we would do th following:

```{r}

# To find the cutting point q, such that p*100 
# percent of the data is less that q

# use qnorm(p,mean,sd)


ans<-round(qnorm(.95,0,1),5)

print(paste('the 95th percentil is ',ans))
```

### Assignment 2

Using what we just learned complete the following:

1.  Find the probability that a normal random variable with mean 68 and standard deviation 2.5 is between 64 and 70.

```{r}

mu <- 68
sigma <- 2.5

ans<-round(pnorm(64,mu,sigma)-pnorm(70,mu,sigma),5)

print(paste(ans))

```

2.  Find a number x0 ,from the standard normal curve, such that P(-x0\<X\<x0)=0.999

```{r}

x0 <- qnorm(0.999/2)

print(x0) #Printing answer


```

Now that we understand the theoretical Normal distribution a little better lets look at taking sample from a normal distribution and seeing how well it compares with the theoretical distribution. To do this we will use rnorm. to take a sample of size 25 from the standard normal distribution we would do the following:

```{r}

# rnorm(size=n,mean,sd) will generate
# a random sample with size = n, from
# a normal distribution with mean and sd


normal_sample_25<-rnorm(25,0,1)

```

We can now plot the histogram and superimpose the theoretical distribution:

```{r}
hist(normal_sample_25,freq = FALSE,
     main = 'Normal Distribution with Sample size 25',
     xlim=c(min(normal_sample_25)-1,max(normal_sample_25)+1))

x<-seq(min(normal_sample_25),max(normal_sample_25),length=200)

y<-dnorm(x,0,1)

lines(x,y)
```

We will put this together in a function that you can use to finish the assignment below

```{r}
normal_dist_plot<-function(n,mu,sigma){
  normal_sample<-rnorm(n,mu,sigma)
  hist(normal_sample,freq=FALSE,main=paste('Normal dsitribution with sample size',n),xlim=c(min(normal_sample)-1,max(normal_sample)+1))
  x<-seq(min(normal_sample),max(normal_sample),length=2000)
  y<-dnorm(x,mu,sigma)
  lines(x,y)
}

normal_dist_plot(200,0,1)


```

3.  Now do the same thing but take samples of n= 50, 100, 1,000, and 10,000

```{r}

#Creating a list of samples that we can iterate through in a for loop
samples = c(10, 100, 1000, 10000)

#Iterate through the list
for (a in samples) {
  normal_dist_plot(a, 0, 1)
}


#Declaring a function to create the normal distribution plot
normal_dist_plot<-function(n,mu,sigma){
  normal_sample<-rnorm(n,mu,sigma)
  hist(normal_sample,freq=FALSE,main=paste('Normal dsitribution with sample size',n),xlim=c(min(normal_sample)-1,max(normal_sample)+1))
  x<-seq(min(normal_sample),max(normal_sample),length=2000)
  y<-dnorm(x,mu,sigma)
  lines(x,y)
}
```

**End of Assignment 2**

Compare the histograms of the three samples to the probability density function for the standard normal distribution. Address how well the histograms compare to the probability density function and what seems to happen to the empirical histograms as the sample size increases.

## The Exponential Distribution

The probability density function of an exponential random variable is:

$f(x)= \lambda e^{-\lambda x}$ for $x \geq 0$

Here $E(x)=\frac{1}{\lambda}=Var(x)$

We can still use the similar d,p,q,r for exponential distribution.

dexp(x, lambda) find the exponential curve value at x

pexp(q, lambda) find the area to the left of cutting point q (P[X\<q]) for exponential.

qexp(p, lambda) find the cutting point q where P[X\<q]=p.

rexp(n, lambda) generate a random sample with size n for exponential.

### Assignment 3

1.  Using the above formula for $f(x)$ find $\lim_{x\to\infty} f(x)$
2.  Generate graphs of the probability density function for the exponential distribution with $E(x)=1,5$ and describe how the graph changes

```{r}
#Question 1:
  # Using the above formula, we will get e^(-λx) = 0, making the entire limit equal 0.

#Question 2:
  x <- seq(-.0000001, 20, length=5000)
  #Defining Lambdas
  lambda1 = 1 # 1/1 = 1
  lambda2 = 1/5
  
  #Calculating PDFs
  pdf_lambda1 <- dexp(x, lambda1)
  pdf_lambda2 <- dexp(x, lambda2)
  
  #-Plotting the PDFs-
  #graph for lamda1
  plot(x, pdf_lambda1, type = 'l')
  #graph for lamda2
  plot(x, pdf_lambda2, type = 'l')
  
  
#Describe how the graph changes:
# When E(x) is increased, the spread becomes larger and the peak (y-intercept) becomes smaller. 
```

3.  Find the probability that an exponential random variable with mean 6 is between 5 and 7

```{r}
mu <- 6
lambda <- 1/mu #calculating lamda for functions

#Calculate the probability
prob_x <- (pexp(7, lambda) - pexp(5, lambda))

#display the probability
print(prob_x)
```

4.Select random samples of size 50, 100 , 1000, and 10,000 from an exponential distribution with mean 1. You can create a function the same way as shown for the normal distribution, but this time use rexp() instead of rnorm().

```{r}
exponential_dist_plot<-function(n,lambda){
  exp_sample<-rexp(n,1/lambda)
  hist(exp_sample,freq=FALSE,main=paste('Exponential dsitribution with sample size',n),xlim=c(0,max(exp_sample)+1))
  x<-seq(min(exp_sample),max(exp_sample),length=2000)
  y<-dexp(x,1/lambda)
  lines(x,y)
}

sampleSizes <- c(50, 100, 1000, 10000)
mu <- 1

#calling function
for(a in sampleSizes) {
  exponential_dist_plot(a, mu)
}


```

5.  Construct histograms for each of the samples and compare them with the probability density function of the exponential distribution with a mean of 1. How well do the histograms compare and what happens as the sample size increases?

    ```{r}
    #Changing the function to also display the probability density
    exponential_dist_plot<-function(n,lambda){
      exp_sample<-rexp(n,1/lambda)
      hist(exp_sample,freq=FALSE,main=paste('Exponential dsitribution with sample size',n),xlim=c(0,max(exp_sample)+1))
      x<-seq(min(exp_sample),max(exp_sample),length=2000)
      y<-dexp(x,1/lambda)
      lines(x,y)
      #Adding the PDF to the histograms
      x <- seq(0, max(exp_sample), length = 2000)
      y <- dexp(x, 1/lambda)
      lines(x, y, col = 'red', lwd = 2)
    }

    #Creating histograms with mu = 1
    n<-c(50,100,1000,10000)
    mu <- 1
    lambda <- mu/1
    for (i in n) {
      exponential_dist_plot(i,lambda)
    }

    #How well does the histogram compare?
    # The PDF and histograms are very similar, reflecting an accurate line of best fit. As the sample size increases, the PDF becomes a better representation of the histogram.
    ```

    **End of Assignment 3**

## Real World Application

### Assignment 4

A. Find three real world examples of normally distributed data that you are interested in. And explain the meaning of pnorm value and qnorm value in the context of your examples.

```         
1) Test Scores
  Scores on standardized tests, such as the SAT, are designed to
  follow a normal distribution. In this contect, the pnorm value 
  would represent the probability that a student scores less than   
  a certain score. For instance, pnorm of 1200 would give the
  probability that an average test-taker would score below an
  1200. Additionally, a pnorm of 1200 - pnorm of 800 would give 
  the probability the average test-taker would score between a
  800 and 1200. The qnorm would instead return percentiles, with 
  a qnorm of .7 representing the score corresponding to the 70th
  percentile.
  
2) Gas Milage in Cars
  Gas milage on cars could definitely reflect a normal
  distribution, with different kinds of cars (sport, hybrid, etc)
  reflecting different ends of gas milage. In this context, pnorm
  would represent the probability that a car would have a lower 
  gas milage that a certain number, such as a pnorm of 30
  representing cars with a gas milage below 30. Additionally,
  qnorm would represent how cars rank in gas milage, with a qnorm
  of .5 representing the 50th percentile of gas milage numbers.
  
3) Household Incomes
  Simlarly to the previous two examples, the economy is set up to
  have a roughly normal distribution curve for income. For this 
  example, the pnorm would represent the probability that a given
  household would have at least that income, with the qnorm 
  representing the household income corresponding to the given 
  quartile.
```

B. Do the same thing as above but for an exponential distribution.

```         
1) Network Times
  The time it takes for a network to retrieve information from a
  server likely follows an exponential distribution. In this
  case, pexp would correspond to to the probability that the 
  time it takes to load a specific websitesomething is less than 
  a certain value. qexp would find the time in which a certain 
  percentile of requests returned in.
  
2) Time between Successive Sales at a Grocery Store
  The time it takes between sales at a grocery store would likely follow an           exponential distrubiton, with most sales happening during rush hours. In this case
  pexp would represent the probability that a sale is made less than a given time.    qexp would represent the time in which successive sales are made given a            percentile. For example, .7 would represent the sales time for the 70th percentile
  
3) Duration of Phone Conversations
  The duration of cell phone conversations would also likely 
  follow an exponential distribution, with most calls lasting 
  shorts amounts of time with longer calls being more rare. In
  this case, pexp would represent the probability that a phone
  call is less than a given duration, while qexp would represent
  the length of the given 
```

**End of Assignment 4**
