---
title: 'Lab 8: Comparing Two Population Means'
author: "kgteller"
date: "2023-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(BSDA)
library(nortest)
```

# Lab 8: Comparing Two Population Means

Today we are going to look at conducting hypothesis tests when we have two samples of data. Recall the steps we want to take when running a hypothesis test:

1.  A description the null and alternative hypotheses in terms of the problem.

2.  A statement of the null and alternative hypotheses in standard symbols.

3.  The significance level (size of the Type I error), α, that will be tolerated.

4.  The name of the statistical test to be used and the assumptions required in order for the test to be valid. (i.e. small sample size, not approximately normal, paired, independent, etc.)

5.  A clear description of exactly what the decision rule is; i.e. under what circumstances is the null hypothesis rejected.

6.  A clear statement indicating whether or not you actually reject H0. This decision must be linked to the numerical value of p actually observed and the value of α.

7.  A very clear description of what was actually decided in the context of the original problem. For example, if the problem is about mean airfares between two cities, simply saying reject Ho is not good enough; you need to say what this decision means as it applies to mean airfares.

8.  Any statistical decision method is subject to Type I or Type II error. Discuss which type is possible in your final decision and if possible, its probability.

When we have two samples we need to first consider the relationship between the data. There are two different relationships we can have between the data:

**Independent Samples**: A random sample of size $n_1$ is selected from the first population, and a random sample of size $n_2$ is selected *independently* from the second population. The null hypothesis usually is $\mu_1=\mu_2$ .Then $\bar{x}_1-\bar{x}_2$is calculated, but this time a *2-sample test* is used; there is no way to view the data as a single sample.

When the sample sizes are small and the the both populations are not normal a Mann-Whitney test can sometimes be used. The Mann-Whitney test would be used if the two samples have a distribution with the same shape. in addition the Mann-Whitney test is not about $\mu_1$ and $\mu_2$, but about the population medians $\eta_1$ and $\eta_2$. so the two hypotheses become:

$H_0: \eta_1=\eta_2 \\$ $H_1: \eta_1 \neq,\lt,\gt \eta_2$

**Paired Samples** When the two samples have the same sample size and each data point in one sample can be *paired* with another data point in the other sample. The difference between the populations means is denoted $\mu_d=\mu_1-\mu_2$. Here the null hypothesis is that $\mu_d=0$. Now this can be treated as a single sample and a 1-sample hypothesis can be run on the **sample differences**.

In order to run either of these tests we can use the t.test(), or wilcox.test() commands that we discussed in the previous lab.

To run an independent 2-sample t-test we would use the following syntax: t.test(sample1,sample2)

to run an independent Mann-whitney test we would use the following syntax: wilcox.test(sample1,sample2,alternative='two.sided','less','greater')

To run a paired t-test we use the following syntax: t.test(sample1,sample2,paired=TRUE)

For today's lab we will use data from the pulse.csv. The data relates to a class of statistic students that measured their resting pulse (Pulse1) and then a random number of students ran in place while the others just stayed still. The pulses of the entire class were recorded again (Pulse2). Along with the pulse data other data was collected that included, height, weight, sex, and whether or not they smoked.

For reference here is a more detailed summary of the data:

In this data set, there are 8 variables:

-   Pulse1: The resting pulse of each student
-   Pulse2: The pulse after running or not running for each student (after treatment)
-   Ran: Whether or not the student ran in place: 1=Yes or 2=No (treatment)
-   Smokes: Whether or not the student smokes regularly: 1=Yes or 2=No
-   Sex: The biological characteristics of males or females of the student: 1=M or 2=F
-   Height: The height of the student, in inches
-   Weight: The weight of the student, in pounds
-   Activity: The usual activity level of the student: 1=Slight, 2=Moderate, or 3=A lot

## Assignment 1

create a data-frame and then create the following objects to answer each of the other assignments:

|     Name      |                    Explanation of Name                    |
|:-------------:|:---------------------------------------------------------:|
|     Diffs     |                       Pulse1-Pulse2                       |
|    Diffs_R    |          Pulse1 -- Pulse2 for Runners (Ran = 1)           |
|  Diffs\_\~R   |        Pulse1 -- Pulse2 for non Runners (Ran = 2)         |
|   Diffs_R&S   |   Pulse1 -- Pulse 2 for Runners who smoke (Smokes = 1)    |
|  Diffs_R&\~S  | Pulse1 -- Pulse2 for Runners who don't smoke (Smokes = 2) |
|   Weight_M    |                     Weights of males                      |
|   Weight_F    |                    Weights of females                     |
|  Weight_M_S   |                  Weights of male smokers                  |
| Weight_M\_\~S |                Weights of male non-smokers                |
|  Weight_F_S   |                 Weights of female smokers                 |
| Weight_F\_\~S |               Weights of female non-smokers               |

```{r}
#Holding the data
data <- read.csv("Pulse.csv")

#Pulse1 - Pulse2
Diffs <- data$Pulse1 - data$Pulse2

#Pulse1 - Pulse2 for Runners
Diffs_R <- data$Pulse1[data$Ran == 1] - data$Pulse2[data$Ran == 1]

#Pulse1 - Pulse2 for Non-Runners
Diffs_nR <- data$Pulse1[data$Ran == 2] - data$Pulse2[data$Ran == 2]

#Pulse1 - Pulse2 for Runners and Smokers
Diffs_RandS <- data$Pulse1[data$Ran == 1 & data$Smokes == 1] - data$Pulse2[data$Ran == 1 & data$Smokes == 1]

#Pulse1 - Pulse2 for Runners and Non-Smokers
Diffs_RandnS <- data$Pulse1[data$Ran == 1 & data$Smokes == 2] - data$Pulse2[data$Ran == 1 & data$Smokes == 2]

#Weights for males
Weight_M <- data$Weight[data$Sex == 1]

#Weight for females
Weight_F <- data$Weight[data$Sex == 2]

#Weight of male smokers
Weight_M_S <- data$Weight[data$Sex == 1 & data$Smokes == 1]

#Weight for male non-smokers
Weight_M_nS <- data$Weight[data$Sex == 1 & data$Smokes == 2]

#Weight of female smokers
Weight_F_S <- data$Weight[data$Sex == 2 & data$Smokes == 1]

#Weight for female non-smokers
Weight_F_nS <- data$Weight[data$Sex == 2 & data$Smokes == 2]
```

**End Assignment 1**

## Assignment 2

A researcher wants to test the hypothesis that the average pulse of the class was greater the second time around than the first time. Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above.

```{r}
#We are comparing averages, so a t-test is to be used
results1 <- t.test(data$Pulse1, data$Pulse2, paired = "true", alternative = "less")
print(results1)
```

```         
Null Hypothesis- The average pulse of the class was equal the first and second time (mu1=mu2)
Alternative Hypothesis- The average pulse of the class before the expirement is less than the class's pulse after the expirement (mu1 < mu2)

α = .05

The sample size is 92 and we are concerned about the means, so a two-sample t-test is best in this scenario

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .000002023, which is significantly smaller than α (.05). Therefore, the Null Hypothesis is rejected.

Since the Null Hypothesis is rejected, we can conclude that the two samples are not equal. Additionally, we can also conclude that the Alternative hypothesis is true, so we know that the mean pulse after treatment is greater than the mean pulse prior to treatment.

There is a 5% probability of a Type 1 error.
```

**End Assignment 2**

Now it probably makes sense to separate the class into two groups, the runners and the non-runners.

## Assignment 3

A researcher now wants to look at the pulses of the running group before and after to conclude whether or not running in place increases mean heart rate. Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above.

```{r}
#Getting the sample size
print(length(Diffs_R))
#We are comparing means, so a t-test is to be used
results2 <- t.test(Diffs_R, alternative = "less")
print(results2)
```

```         
Null Hypothesis: The mean pulses of indivuals receiving treatment is the same before and after
ALternative Hypothesis: The mean pulses of individuals receiving treatment is greater after treatment compared to before treatment

α = .05

The sample size is 35 (large enough) and we are concerned about the means, so a one-sample t-test is best in this scenario

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .000000005358, which is significantly smaller than α (.05). Therefore, the Null Hypothesis is rejected.

Since the Null Hypothesis is rejected, we can conclude that subjects do not have the same pulse before and after treatment. Additionally, we can also conclude that the Alternative hypothesis is true, so we know that subjects have a higher pulse after treatment when compared to before treatment.

There is a 5% probability of a Type 1 error.
```

**End Assignment 3**

## Assignment 4

A researcher now wants to run the same hypothesis as before but with the non-runners to see if there is any difference in mean heart rate. Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above.

```{r}
#Getting the sample size
print(length(Diffs_nR))
#We are comparing means and the population is large enough, so a t-test is to be used
results3 <- t.test(Diffs_nR, alternative = "less")
print(results3)
```

```         
Null Hypothesis: The mean pulses of non-runners (people not receiving treatment) and is the same before and after treatment
ALternative Hypothesis: The mean pulses of non-runners (people not receiving treatment) is greater after when compared to before

α = .05

The sample size is 57 and we are concerned about the means, so a one-sample t-test is best in this scenario

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .5754, which is larger than α (.05). Therefore, the Null Hypothesis is not rejected.

The Null Hypothesis is not rejected, so we can assume that the mean pulses for non-runners before and after treatment is the same.

There is a 5% probability of a Type 1 error.
```

**End Assignment 4**

## Assignment 5

Now among the runners there could be another factor that affects the average heart rate. This confounding factor could be smoking. Therefore the researcher wants to know does the mean heart rate of smokers increase after running in place? Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above. You may have a small sample size and it is non-normal you may have to use the Wilcoxon or Sign tests discussed in the previous lab.

```{r}
#Getting the sample size
print(length(Diffs_RandS))
#Since the sample size is small, testing for normality
ad.test(Diffs_RandS)
#Population is normal, so we run a tTest
results4 <- t.test(Diffs_RandS, alternative = "less")
print(results4)
```

```         
Null Hypothesis: The mean heart rates of runners (people who receive treatment) who smoke is the same before and after.
ALternative Hypothesis: The mean pulses of non-runners (people not receiving treatment) who smoke is greater after when compared to before.

α = .05

The sample size is 92 and we are concerned about the means, so a one-sample t-test is best in this scenario

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .0008227, which is smaller than α (.05). Therefore, the Null Hypothesis is rejected.

The Null Hypothesis is rejected so we can assume that smoking subjects do not have the same heart rate before running when compared to after. Additionally, We can assume that Alternative Hypothesis that smokers have a higher heart rate after running when compared to before.

There is a 5% probability of a Type 1 error.
```

**End Assignment 5**

## Assignment 6

Now to make a comparison the next hypothesis would be does running in place increase the mean heart rate of non-smokers? Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above.

```{r}
#Getting the sample size
print(length(Diffs_RandnS))
#Since the sample size is small, testing for normality
ad.test(Diffs_RandnS)
#Population is normal, so we run a t-test
results5 <- t.test(Diffs_RandnS, alternative="less")
print(results5)
```

```         
Null Hypothesis: The mean heart rates of runners (people who receive treatment) who do not smoke is the same before and after.
ALternative Hypothesis: The mean pulses of non-runners (people not receiving treatment) who do not smoke is greater after when compared to before.

α = .05

The sample size is 23 and we are concerned about the means. Additionally, the data is normal. Therefore, a T-test is best in this scenario.

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .000001113, which is smaller than α (.05). Therefore, the Null Hypothesis is rejected.

The Null Hypothesis is rejected so we can assume that non-smoking subjects do not have the same heart rate before running when compared to after. Additionally, We can assume that Alternative Hypothesis that non-smokers have a higher heart rate after running when compared to before.

There is a 5% probability of a Type 1 error.
```

**End Assignment 6**

## Assignment 7

Now a researcher is interested in the weights of the class. They want to determine whether the weights of males and females differ.Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above. ( if using a non-parametric test remember to state the conclusion in terms of means and not medians)

```{r}
#Getting sizes for test determination
print(length(Weight_M)) #Male 
summary(Weight_M) #Validating p-value
print(length(Weight_F)) #Female
summary(Weight_F) #Validating
#Both sizes are sufficiently large, so running a 2-sample t-test
result7 <- t.test(Weight_M, Weight_F) #default alternative (not equal)
print(result7)
```

```         
Null Hypothesis: There is no difference in the weight of males and females
ALternative Hypothesis: There is a difference in the weight of males and females

α = .05

The sample size for both samples are sufficiently large. Therefore, a T-test is best in this scenario.

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is esentially 0, which is obviously less than α (.05). Therefore, the Null Hypothesis is rejected.

The Null Hypothesis is rejected so we can assume the Alternative Hypothesis, which is that the weight of males and females differ.

There is a 5% probability of a Type 1 error.
```

**End Assignment 7**

## Assignment 8

Going a little deeper into the data a researcher thinks that smoking might have an effect on weight. The researcher wants to test the hypothesis about whether the mean weight of males who smoke is different from males who do not smoke. Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above. ( if using a non-parametric test remember to state the conclusion in terms of means and not medians)

```{r}
#Collecting basic statistics for validating results
summary(Weight_M_S)
print(length(Weight_M_S))
summary(Weight_M_nS)
print(length(Weight_M_nS))
#Getting normality, since both samples are not sufficiently large
ad.test(Weight_M_S)
ad.test(Weight_M_nS)
#Running appropriate test
result8 <- t.test(Weight_M_S, Weight_M_nS) #Default Alternative
print(result8)
```

```         
Null Hypothesis: There is no difference in the weight of males who smoke and males who do not smoke
ALternative Hypothesis: There is a difference in the weight of males who smoke and males who do not smoke

α = .05

The sample size of male smokers is not sufficiently large (20), however the distribution is norma. Additionally, the sample size of male non-smokers is sufficiently large. Therefore, a wilcox test is best to use.

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .1834, which is greater than α (.05). Therefore, the Null Hypothesis is not rejected.

The Null Hypothesis is not rejected, so we can assume that smoking does not affect male weight.

There is a 5% probability of a Type 1 error.
```

**End Assignment 8**

## Assignment 9

The last hypothesis that the researcher want to test is whether the mean weight of females who smoke is different from females who do not smoke. Conduct the appropriate test to test this hypothesis and clearly state your conclusion following the 8 steps described above. ( if using a non-parametric test remember to state the conclusion in terms of means and not medians)

```{r}
#Collecting basic statistics for validating results
summary(Weight_F_S)
print(length(Weight_F_S))
summary(Weight_F_nS)
print(length(Weight_F_nS))
#Getting normality, since both samples are not sufficiently large
ad.test(Weight_F_S)
ad.test(Weight_F_nS)
#Running appropriate test
result9 <- t.test(Weight_F_S, Weight_F_nS) #Default Alternative
print(result9)
```

```         
Null Hypothesis: There is no difference in the weight of females who smoke and males who do not smoke
ALternative Hypothesis: There is a difference in the weight of females who smoke and males who do not smoke

α = .05

Neither sample size is sufficiently large (8 & 27). However, both samples are normal, so a t-test is appropriate in this scenario.

If the p-value is less than α (.05) we reject the Null Hypothesis

The p-value of this test is .5281, which is greater than α (.05). Therefore, the Null Hypothesis is not rejected.

The Null Hypothesis is not rejected, so we can assume that smoking does not affect female weight.

There is a 5% probability of a Type 1 error.
```

**End Assignment 9**

## Conclusion assignment

Based on the hypotheses you have tested and there results write a concluding paragraph or more about your findings:

```         
This expirement features 92 participants, and aims to measure their heartrate before and after "treatment" (running in place). 35 participants received treatment, and 57 paricipants were used as a control group and did not receive treatment. Additionally, several characteristics were noted, such as their weight, gender, and if they smoked.
Through hypothesis testing, numerous conclusions can be made the population of which this expirement tested. Firstly, we can assume that receiving treatment (running in place) will affect heart rate. Through running the appropriate tests, it was determined that running in place increased heart rate while not running in place did not increase heart rate. Additionally, the tests did not prove that the factor of smoking has an effect on heart rate before and after treatment, with both smokers and non-smokers experiencing an increased heart rate after running in place. Additionally, the tests state that there is an enormously high chance that male weight is different than male weight. Lastly, the study and hypothesis testing wanted to examine if smoking impacts male or females weight, in which is concluded that neither male nor female weight is affected by smoking. 
```

**End Conclusion Assignment**
